---
title: "Description and usage of MsBackendTimsTof"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Description and usage of MsBackendTimsTof}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{MsBackendTimsTof}
    %\VignetteDepends{Spectra,BiocStyle,BiocParallel}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
BiocStyle::markdown()
```

**Package**: `r Biocpkg("MsBackendTimsTof")`<br />
**Authors**: `r packageDescription("MsBackendTimsTof")[["Author"]] `<br />
**Compiled**: `r date()`

```{r, echo = FALSE, message = FALSE}
library(Spectra)
library(BiocStyle)
```

# Introduction

The `r Biocpkg("Spectra")` package provides a central infrastructure for the
handling of Mass Spectrometry (MS) data. The package supports interchangeable
use of different *backends* to import MS data from a variety of sources (such as
mzML files). The `MsBackendTimsTof` package adds support for Bruker TimsTOF raw
data files. This vignette shows how, and which data can be retrieved from such
files.

# Installation

The package depends on the [OpenTIMS](https://github.com/michalsta/opentims) C++
library to access data in timsTOF Pro data format (TDF) which is provided by the
`opentimsr` R package. The `MsBackendTimsTof` package can be installed with:

```{r, eval = FALSE}
BiocManager::install("RforMassSpectrometry/MsBackendTimsTof")
```

To get some variables from the data files an additional library from the
manufacturer is needed. This library can be downloaded with:

```{r, eval = FALSE}
so_folder <- tempdir()
library(opentimsr)
so_file <- download_bruker_proprietary_code(so_folder)
```

This downloads the shared library to a temporary folder. Note however that at
present this shared library is **only available for Windows and Linux** (i.e. no
macOS support). Next, to use this library, it has to be registered with the
`opentimsr` package:

```{r, eval = FALSE}
setup_bruker_so(so_file)
```

These steps would be necessary for every new R session. To avoid that, it is
suggested to copy the downloaded shared library above to a directory on the
computer and to define an environment variable called `TIMSTOF_LIB` that defines
the full path where this file is located (i.e. a character string defining the
full file path with the file name). This variable can either be defined system
wide, or within the *.Rprofile* file. An example entry in a *.Rprofile* could
for example be:

```
options(TIMSTOF_LIB = "/Users/jo/lib/libtimsdata.so")
```


# Accessing data from Bruker TimsTOF files

The `MsBackendTimsTof` package adds support for Bruker TimsTOF files to
`Spectra`-based analysis workflows. Below we load the package and in addition
fetch the required shared library and store that to a temporary folder.

```{r, message = FALSE}
library(MsBackendTimsTof)

## Load the opentimsr package and download and register the shared library
library(opentimsr)
so_folder <- tempdir()
so_file <- download_bruker_proprietary_code(so_folder, method = "wget")
setup_bruker_so(so_file)
```

As detailed in the installation section, the code to download the shared library
would only be necessary once, if the path to this file is defined in a
environment variable `TIMSTOF_LIB`.

We next load the TDF test file which is bundled within this package.

```{r}
fl <- system.file("ddaPASEF.d", package = "MsBackendTimsTof")

be <- backendInitialize(MsBackendTimsTof(), fl)
```

In a real use case, we would however directly load the data into a `Spectra`
object:

```{r}
sps <- Spectra(fl, source = MsBackendTimsTof())
sps
```

We thus have access to all spectra variables within the file:

```{r}
spectraVariables(be)
```

And the full data can be retrieved with `spectraData`:

```{r}
all <- spectraData(be)
```

The data is organized by individual spectra, all spectra measured within the
same *frame* have the same value in the spectra variable `"frameId"`. `"tof"`
and `"inv_ion_mobility"` are available for each individual peak (not spectrum)
and are thus additional *peaks variables*.

Below we subset the backend to a range of spectra and extract their `peaksData`.

```{r}
be_sub <- be[218:226]
peaksData(be_sub, columns = c("mz", "intensity", "tof", "inv_ion_mobility",
                              "retention_time"))
```

Open questions:

- is the data representation OK? Does it make sense to have `inv_ion_mobility`
  as peaks variable, or should that be a spectra variable (seems to be the same
  value for all peaks of one spectrum). Same for `retention_time`.

```{r}
## Check if retention time is the same for all peaks within a spectrum
pks <- peaksData(be, columns = c("mz", "intensity", "tof", "inv_ion_mobility",
                                 "retention_time"))
rt <- vapply(pks, function(z) {
    if (nrow(z)) {
        if (length(unique(z[, "retention_time"])) != 1L)
            stop("different values for retention_time")
        z[1, "retention_time"]
    } else NA_real_
}, numeric(1))
iim <- vapply(pks, function(z) {
    if (nrow(z)) {
        if (length(unique(z[, "inv_ion_mobility"])) != 1L)
            stop("different values for inv_ion_mobility")
        z[1, "inv_ion_mobility"]
    } else NA_real_
}, numeric(1))

## Check if the retention times from peaks table match the ones from spectra
## variables.
all.equal(all$rtime[!is.na(rt)], rt[!is.na(rt)])

```

So, summarizing: seems the retention time information in the peaks data is
redundant and also the inverse ion mobility seems to be rather a spectra, not a
peaks variable.

# Session information

```{r}
sessionInfo()
```
