---
title: "Description and usage of *MsBackendTimsTof*"
output:
    BiocStyle::html_document:
        toc_float: true
vignette: >
    %\VignetteIndexEntry{Description and usage of MsBackendTimsTof}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
    %\VignettePackage{MsBackendTimsTof}
    %\VignetteDepends{Spectra,BiocStyle,BiocParallel}
---

```{r style, echo = FALSE, results = 'asis', message=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
library(BiocStyle)
BiocStyle::markdown()
```

# Introduction

The `r Biocpkg("Spectra")` package provides a central infrastructure for Mass
Spectrometry (MS) data in R. The package supports interchangeable use of
different *backends* to represent or import MS data from a variety of formats
and sources (such as mzML files). The *MsBackendTimsTof"* package adds support
for MS data from Bruker TimsTOF raw data files. This vignette shows how, and
which data can be retrieved from such files.

# Installation

The package depends on the [OpenTIMS](https://github.com/michalsta/opentims) C++
library to access data in timsTOF Pro data format (TDF) which is provided by the
*opentimsr* R package. The *opentimsr* and *MsBackendTimsTof* package can be
installed with:

```{r installation, eval = FALSE}
BiocManager::install("michalsta/opentims/src/opentimsr")
BiocManager::install("RforMassSpectrometry/MsBackendTimsTof")
```

Some variables in the data files can only be accessed after setting up a
converter library. Two options are available.
The first is to use the open-source library built into *opentimsr*:

```{r setup-opensource, eval = FALSE}
library(MsBackendTimsTof)

setup_converter_library()
```

The `setup_converter_library()` function is also run on package startup hence,
to use the default, no manual call to `setup_converter_library()` is needed.

Alternatively it is possible to use the proprietary Bruker converter library,
which can be downloaded and cached through *BiocFileCache* or downloaded to a
personal folder. Note that this library is currently **only available for
Windows and Linux** (no macOS support).

```{r setup-bruker, eval = FALSE}
setup_converter_library(opensource = FALSE)
```


# Accessing data from Bruker TimsTOF files

The `r Biocpkg("MsBackendTimsTof")` package adds support for Bruker TimsTOF
files to `Spectra`-based analysis workflows. Below we load the package which
will also automatically setup the use of the open-source converter library.

```{r load-library, message = FALSE}
library(MsBackendTimsTof)
```

As detailed in the installation section, the code to download the shared library
would only be necessary once, if the path to this file is defined in a
environment variable `TIMSTOF_LIB`.

We next load the TDF test file which is bundled within this package.

```{r initialize}
## Get the path/file name of the example file
fl <- system.file("ddaPASEF.d", package = "MsBackendTimsTof")

## Initialize the MsBackend
be <- backendInitialize(MsBackendTimsTof(), fl)
be
```

In a real use case, we would however directly load the data into a `Spectra`
object:

```{r load-spectra}
sps <- Spectra(fl, source = MsBackendTimsTof())
sps
```

We thus have access to all spectra variables within the file:

```{r spectra-variables}
spectraVariables(sps)
```

And the full data for spectra variables can be retrieved with `spectraData`:

```{r full-spectra-data}
all <- spectraData(sps)
all
```

The data is organized by individual spectra, all spectra measured within the
same *frame* have the same value in the spectra variable `"frameId"`. This also
means that the `scanIndex` is **not** unique within a file, only within one
*frame*:

```{r duplicated-scan}
sps$scanIndex |> anyDuplicated()
```

The combination of `frameId` and `scanIndex` identifies each spectrum uniquely
within the same file:

```{r frame-scan}
spectraData(sps, c("frameId", "scanIndex", "file"))
```

The *inverse ion mobility* data is provided by the `"inv_ion_mobility"`
variable. This information can be extracted as a *spectra variable* (i.e.,
one value per spectrum):

```{r ion-mobility-spectra-variable}
## Get the ion mobility information
sps$inv_ion_mobility |> head()
```

But can also be requested to be added to the `peaksData()`. Below we extract the
full peaks data (i.e., all *peaks variables* for all spectra) and show the
information of one selected spectrum:

```{r peaks-data}
## Request all supported peaks variables
pd <- peaksData(sps, peaksVariables(sps))
## Show the data for the spectrum number 10
pd[[10L]]
```

Note however that `"inv_ion_mobility"` and `"retention_time"` have the
same value for all peaks in each spectrum. Thus, these variables should be
accessed not through `peaksData()`, but through `spectraData()` or using the `$`
operator or the dedicated `rtime()` function.

The `MsBackendTimsTof` is a *on-disk* backend, hence reading peaks and spectra
variables from the original data file(s) upon request. Modifying spectra
variables within the raw data files is not possible (and also not desired), but,
by directly extending the `MsBackendCached` backend from the
`r Biocpkg("Spectra")` package `MsBackendTimsTof` supports changing or adding
spectra variables by *caching* them locally within the object. Below we add a
new spectra variable to the object.

```{r add-variable}
## Add a new spectra variable
sps$new_variable <- seq_along(sps)
```

This new variable is now available within the object and can be extract like
any other spectra variable.

```{r show-variables}
spectraVariables(sps)
sps$new_variable |> head()
```

Analogously it is also possible to change existing spectra variables. Below we
shift each retention time by 10 seconds.

```{r shift-rtime}
rtime(sps) |> head()
sps$rtime <- rtime(sps) + 10
rtime(sps) |> head()
```


# Session information

```{r}
sessionInfo()
```
